---
title: "Project Assignment 1"
author: "Nannette Orsini Spear"
date: "April 29, 2018"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

#Loading and pre-processing the data.

Load knitr package
```{r load knitr, echo=TRUE}
library(knitr)
```

Set the Working Directory.
```{r set working directory, echo=TRUE}
opts_knit$set(root.dir = "C:/RStudio/Coursera/Reproducible Research")
```

Read the dataset.
```{r read data, echo=TRUE}
activity <- read.csv("C:/RStudio/Coursera/Reproducible Research/Data Files/activity.csv",header = TRUE, sep = ",",stringsAsFactors = FALSE)
str(activity)
``` 

Change date field from character to date.
```{r date, echo=TRUE}
activity$date <- as.Date(activity$date)
class(activity$date)
```

#What is the mean total number of steps taken per day?

Total number of steps each day.
```{r mean, echo=TRUE}
dailysteps <- aggregate(steps~date,activity,sum)
dailysteps
```

Histogram of the total number of steps taken each day.
```{r dailysteps, echo=TRUE}
hist(dailysteps$steps,
     xlab = "Total Steps per Day",
     ylab = "Frequency of Days",
     main = "Histogram of Total Number of Steps per Day",
     col = "steelblue")
```

Mean number of steps taken each day.
```{r dailymean, echo=TRUE}
dailymean <- aggregate(steps~date,activity,mean)
dailymean
```

Median number of steps taken each day.
```{r dailymedian, echo=TRUE}
dailymedian <- aggregate(steps~date,activity,median)
dailymedian
```

#What is the average daily activity pattern?

The average number of steps taken for each interval.
```{r intervalmean, echo=TRUE}
intervalmean <- aggregate(steps~interval, activity, mean)
intervalmean
```

Each day has 288, five minute intervals.  The plot below represents the average number of steps for each interval across all days.
```{r plot intervalmean, echo=TRUE}
plot(intervalmean$interval, intervalmean$steps, type = "l", xlab = "5 Minute Intervals",
     ylab = "Average Number of Steps", main = "Average Number of Steps per Interval",
     col = "dark green", lwd = 1)
```

The 5-minute interval, on average and across all days, contains the maximum number of steps.
```{r max intervalmean, echo=TRUE}
intervalmean[which.max(intervalmean$steps),]
```

#Imputing Missing Values

Total number of rows with NAs.
```{r na, echo=TRUE}
sum(is.na(activity$steps))
```

Use the average steps over all intervals and across all days in place of the missing values.
```{r activitymean, echo=TRUE}
activitymean <- mean(activity$steps,na.rm = TRUE)
activitymean
```

New data set to impute missing values.
```{r new data set, echo=TRUE}
activity2 <- activity
```

#Imputing missing values:

Assigning mean to 10/1/2012
```{r Oct 1, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-10-01",])
activity2[is.na(activity2$steps) & activity2$date == "2012-10-01", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-10-01",])
```

Assigning mean to 10/8/2012
```{r Oct 8, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-10-08",])
activity2[is.na(activity2$steps) & activity2$date == "2012-10-08", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-10-08",])
```

Assigning mean to 11/1/2012
```{r Nov 1, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-01",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-01", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-01",])
```

Assigning mean to 11/4/2012
```{r Nov 4, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-04",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-04", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-04",])
```

Assigning mean to 11/9/2012
```{r Nov 9, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-09",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-09", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-09",])
```

Assigning mean to 11/10/2012
```{r Nov 10, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-10",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-10", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-10",])
```

Assigning mean to 11/14/2012
```{r Nov 14, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-14",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-14", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-14",])
```

Assigning mean to 11/30/2012
```{r Nov 30, echo=TRUE}
summary(activity2[is.na(activity2$steps) & activity2$date == "2012-11-30",])
activity2[is.na(activity2$steps) & activity2$date == "2012-11-30", "steps"] <- activitymean
summary(activity2[activity2$date=="2012-11-30",])
```

Total number of rows with NAs
```{r remaining na, echo=TRUE}
sum(is.na(activity2$steps))
```

Total number of steps each day
```{r new daily steps na, echo=TRUE}
dailysteps2 <- aggregate(steps~date,activity2,sum)
dailysteps2
```

New Histogram of the total number of steps taken each day with the overall mean of dataset replacing the missing values.
```{r new hist, echo=TRUE}
library(ggplot2)

hist(dailysteps2$steps,
     xlab = "Total Steps per Day",
     ylab = "Frequency of Days",
     main = "Histogram of Total Number of Steps per Day",
     col = "steelblue")
```

Comparison of the total number of steps per day.  Replacing the missing values with the overall mean increases the number of days having 10,000 to 15,000 steps per day.
```{r comparison, echo=TRUE}
par(mfrow = c(1,2), mar = c(4,4,2,1), oma = c(0,0,2,0))
hist(dailysteps$steps,
     xlab = "Total Steps per Day",
     ylab = "Frequency of Days",
     main = "Missing Values",
     col = "steelblue")

hist(dailysteps2$steps,
     xlab = "Total Steps per Day",
     ylab = "Frequency of Days",
     main = "Mean Replacing Missing Values",
     col = "steelblue")
```

Updated Mean number of steps taken each day
```{r updated mean, echo=TRUE}
dailymean2 <- aggregate(steps~date,activity2,mean)
dailymean2
```

Updated Median number of steps taken each day
```{r updated median, echo=TRUE}
dailymedian2 <- aggregate(steps~date,activity2,median)
dailymedian2
```
##Are there differences in activity patterns between weekdays and weekends?
###Yes, weekends have more steps throughout the day compared to weekdays.
###During the weekday, the average steps are highest in the early part of the day but drop significantly during the remainder of the day.  
###During the weekends, activity appears more consistent throughout most of the day.

Create new factor variables:  "weekday" and "weekend"
```{r new variable, echo=TRUE}
activity2$week <- ifelse(weekdays(activity2$date)%in% c("Saturday","Sunday"),"weekend", "weekday")
activity2$week <- factor(activity2$week)
str(activity2)
```
Average steps by interval across all weekdays and weekends
```{r week, echo=TRUE}
intervalmean2 <- aggregate(steps~interval + week, activity2, mean)
intervalmean2
```

Plotting average steps by interval across Weekdays and Weekends
```{r panel plot, echo=TRUE}
library(ggplot2)
ggplot(intervalmean2, aes(interval, steps)) +
      geom_line(color = "blue", linetype = 1) + 
      geom_smooth(method = "lm", size = 1, linetype = 3, color = "black") +
      ggtitle(paste("Average Number of Steps Taken per Interval", "\n", "Across Weekdays and Weekends")) +
      facet_wrap(~week)
```